---
title: "Cause and Effect"
author: "Yubo Cao(Netid:yubocao2)"
date: "11/1/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/markcao/Documents/Course/2019Fall/Research/top100")
options(scipen = 999)
```

# PREPARERATION

## Set up libraries and include tests from AMN

```{r include=FALSE}
library(mboost)
library(parallel)
library(stabs)
library(kernlab)
library(readr)
library(lubridate)
library(dplyr)

dir = "/Users/markcao/Documents/Course/2019Fall/Research/top100"
source("../fitting.R")
source("../indtestHsic.R")
files = list.files(dir)
```

## Helper functions for ANM test

```{r helper}
test <- function(x, y, reg_model = train_GAMboost, ind_test = indtestHsic, cut_off = 10, verbose = FALSE, detail = FALSE) {
    xToY <- reg_model(x, y)
    yToX <- reg_model(y, x)
    xToYMod <- ind_test(xToY$residuals, x)
    yToXMod <- ind_test(yToX$residuals, y)
    xToY.P <- xToYMod$p.value
    yToX.P <- yToXMod$p.value

    if (xToY.P > 10 * yToX.P) {
        result <- TRUE
    } else if (xToY.P * 10 < yToX.P) {
        result <- FALSE
    } else {
        result <- NA
    }

    if (verbose) {
        message("P value for forward is ", xToY.P, ", P value for backword is ", yToX.P, " result is ", ifelse(is.na(result), "Inconclusive", result))
    }

    ans <- c(result, xToY.P, yToX.P)
    if (detail) {
        ans <- c(result, xToY.P, as.vector(xToY$model$coeff), yToX.P, as.vector(yToX$model$coeff))
    }
    
    ans
}

shift <- function(x, col_names, lags = 1) {
    name <- "y"
    x[, name] <- NA
    x[1 : (nrow(x) - lags), name] <- x[(lags + 1) : nrow(x), col_names]
    x = x[1 : (nrow(x) - lags), ]
}

file_test <- function(file, col_names, reg_model = train_GAMboost, ind_test = indtestHsic, cut_off = 10, lags = 1, collection = 1 : 12, verbose = FALSE, detail = FALSE) {
    data <- read.csv(file, skip = 11, stringsAsFactors = FALSE)
    data[data == " M"] <- NA
    data$x <- as.numeric(data[, col_names])
    data <- data[, c("Date", "x")]
    data <- na.omit(shift(data, "x", lags))
    
    collection_result <- matrix(0, length(collection), ifelse(detail, 7, 3))
    for (i in collection) {
        curr <- data[month(data$Date) %in% i, ]
        result <- test(curr$x, curr$y, reg_model = reg_model, verbose = verbose, detail = detail)
        collection_result[i, ] = c(ifelse(is.na(result[1]), -1, as.numeric(result)[1]), result[2: length(result)])
    }

    return(list(collection_result, data))
}

var_test_oneway <- function(data, x = "x", y = "y", quane = seq(0, 1, 0.1)) {
    result <- rep(0, length(quan - 1))
    q <- quantile(data[, x], probs = quan)
    for (i in 1 : (length(result) - 1)) {
        result[i] = var(data[data[, x] >= q[i] & data[, x] < q[i + 1], y], na.rm = TRUE)
    }

    result
}

var_test <- function(data, x = "x", y = "y", quantile = seq(0, 1, 0.1), threshold = 2) {
    xToY <- var(var_test_oneway(data, x, y), na.rm = TRUE)
    yToX <- var(var_test_oneway(data, y, x), na.rm = TRUE)

    if (xToY > threshold * yToX) {
        return("X~Y")
    } else if (yToX > threshold * xToY) {
        return("Y~X")
    } else {
        return("Inconclusive")
    }
}
```

# PROCESSING

## 100 Places weather datasets

```{r eval=FALSE, echo=TRUE}
result_coef <- data.frame(FileName = character(),
                          TestParameter = character(),
                          TestLag = integer(),
                          RegressionModel = character(),
                          Collection = factor(),
                          ForwardPValue = double(),
                          ForwardIntersection = double(),
                          ForwardBeta = double(),
                          BackwardPValue = double(),
                          BackwardIntersection = double(),
                          BackwardBeta = double(),
                          Std = double(),
                          Result = factor())

count <- 0
for (file in files) {
    message("Start ", count, " testing file: ", file)
    count <- count + 1
    f_result <- file_test(file = file,
                          col_names = "Max.Temperature",
                          reg_model = train_linear,
                          lags = 1,
                          detail = TRUE)

    data <- f_result[[2]]
    f_result <- f_result[[1]]
    
    for (r in (1 : nrow(f_result))) {
        curr_result <- factor("TRUE")
        if (f_result[r, 1] == 0 ) {
            curr_result <- factor("FALSE")
        } else if (f_result[r, 1] == -1 ) {
            curr_result <- factor("Inconclusive")
        }

        curr_month <- data[month(data$Date) %in% r, ]
        curr_month$day <- day(curr_month$Date)
        curr_std <- (curr_month %>% group_by(day) %>% summarise(sd = sd(x)) %>% summarise(mean = mean(sd)))$mean
        
        curr <- list(FileName = file,
                     TestParameter = "Max.Temperature",
                     TestLag = 1,
                     RegressionModel = "train_liear",
                     Collection = r,
                     ForwardPValue = f_result[r, 2],
                     ForwardIntersection = f_result[r, 3],
                     ForwardBeta = f_result[r, 4],
                     BackwardPValue = f_result[r, 5],
                     BackwardIntersection = f_result[r, 6],
                     BackwardBeta = f_result[r, 7],
                     Std = curr_std,
                     Result = curr_result)

        result_coef <- rbind(result_coef, curr, stringsAsFactors = FALSE)
    }
}

write.csv(result_coef, "../AllLinear.csv", row.names = FALSE)
```

## Additional data sets: 42, 43, 44 in pairs

```{r}
f42 <- read.delim("../pair0042.txt", header = FALSE, stringsAsFactors = FALSE)
f43 <- read.delim("../pair0043.txt", header = FALSE, sep = "", stringsAsFactors = FALSE)
f44 <- read.delim("../pair0044.txt", header = FALSE, sep = "", stringsAsFactors = FALSE)

f42_result <- test(f42$V1, f42$V2, reg_model = train_linear, detail = TRUE)
f43_result <- test(f43$V1, f43$V2, reg_model = train_linear, detail = TRUE)
f44_result <- test(f44$V1, f44$V2, reg_model = train_linear, detail = TRUE)
```


# ANALYSIS

We first look at the accuracy of the tests in 100 places datasets.

```{r echo=TRUE}
result_coef <- read.csv("../AllLinear.csv", stringsAsFactors = FALSE)
xx <- result_coef

accu <- xx %>% group_by(FileName) %>% filter(Result == "TRUE") %>% count() %>% arrange(desc(n)) %>% ungroup %>% as.data.frame()

hist(accu$n, xlab = "Number of TRUE", main = "Accuracy for each location")
message("Overall accuracy is ", sum(xx$Result == "TRUE") / nrow(xx))
```

The overall accuracy is very good.

Let's take a look at which month has the best performance.

```{r 100}
accu <- xx %>% group_by(Collection) %>% filter(Result == "TRUE") %>% count() %>%  ungroup %>% as.data.frame()

month_accu <- accu$n
names(month_accu) <- accu$Collection
barplot(month_accu, xlab = "Month", ylab = "Accuracy", main = "Accuracy for Each Month", ylim = c(0, 100))
```

We notice that the accuracy is best in summer, then winter, then other seasons.

We check the stability of the test.

In "AR_NEWPORT" Jan is False, Feb is Inconclusive, March is False

```{r}
file = "AR_NEWPORT.csv"
data <- read.csv(file, skip = 11, stringsAsFactors = FALSE)
data[data == " M"] <- NA
data$x <- as.numeric(data[, "Max.Temperature"])
data <- data[, c("Date", "x")]
data_All <- na.omit(shift(data, "x", 1))

accu <- matrix(rep(0, 30 * 3), nrow = 3)
for (i in seq(100, 3000, 100)) {
  
  for (r in 1:3) {
    total <- 0
    data <- data_All[month(data_All$Date) %in% r, ]
    for (j in 1:10) {
      xx <- data[sample(1 : nrow(data), i), ]
      yy <- test(as.numeric(xx$x), as.numeric(xx$y), reg_model = train_linear, detail = TRUE)
      
      curr_result <- factor("X->Y")
      if (is.na(yy[1])) {
          curr_result <- factor("Inconclusive")
      } else if (yy[1] == 0) {
          curr_result <- factor("Y->X")
      }
      
      if (curr_result == "X->Y") {
        total <- total + 1
      }
    }
    
    accu[r, i / 100] = total / 10.0
  }
  
}

for (i in 1 : 3) {
  plot(1:30, accu[i, ], main = paste("Month ", i))
}
```


```{r monthAndParaPlot}
xx <- result_coef
a <- xx %>% group_by(Collection) %>% summarize_at(vars(c(starts_with("Forward"), starts_with("Backward"), "Std")),list(mean = mean, sd = sd)) %>% as.data.frame()

plot(x = 1 : 12, type = "n")
for (i in 1 : 14) {
  points(a[, i + 1], col = rainbow(14)[i], type = "l")
  message("The correlation between Accuracy and ", colnames(a)[i + 1], " is ", cor(month_accu, a[, i + 1]))
}

legend(x = 1, y = 12, legend = colnames(a)[-1], col = rainbow(14), lty = 1, cex = 0.4)
```

```{r seperatePlots}
par(mfrow = c(1, 2))
for (i in 1 : 14) {
  
  plot(a[, i + 1], type = "l", pch = i, main = colnames(a)[i], xlab = "Month", ylab = "Val")
}

par(mfrow = c(1, 1))

```


A regression test on ForwardBeta

```{r regression}
set.seed(541)
whole <- result_coef[, c("FileName", "Collection", "ForwardBeta", "ForwardIntersection", "Std", "Result")]
whole[whole$Result == "Inconclusive" | whole$Result == "FALSE", "Result"] = 0
whole[whole$Result == "TRUE", "Result"] = 1
whole[, "Result"] <- as.numeric(whole[, "Result"])

all_train_ind <- sample(1200, 800)
all_train <- whole[all_train_ind, ]
all_test <- whole[-all_train_ind, ]

accuracy <- rep(0, 10)
model <- list()
for (i in 1 : 10) {
    train_ind <- sample(800, 600)
    train_partial <- all_train[train_ind, ]
    test_partial <- all_train[-train_ind, ]
    m <- glm(Result ~ ForwardBeta + Std + ForwardIntersection, data = train_partial)
    p <- predict(m, test_partial, type = "response")
    r <- ifelse(p > 0.5, 1, 0)
    accuracy[i] <- sum(r == test_partial$Result) / 200.0
    model[[i]] <- m
}

model <- model[[which.max(accuracy)]]
p <- predict(model, all_test, type = "response")
r <- ifelse(p > 0.5, 1, 0)
all_accu <- sum(r == all_test$Result) / 400.0

message("The accuracy is ", all_accu)
```

Now select one file that performs well to do a stable test
```{r stability}
good <- result_coef %>% filter(Result == "TRUE") %>% group_by(FileName) %>% count() %>% arrange(n) %>% filter(n >= 9)
good_files <- good$FileName

file <- good_files[5]
data <- read.csv(file, skip = 11, stringsAsFactors = FALSE)
data[data == " M"] <- NA
data$x <- as.numeric(data[, "Max.Temperature"])
data <- data[, c("Date", "x")]
data <- na.omit(shift(data, "x", 1))

mult <- 1
noise <- seq(0, 10, 1)
temp <- data
acc <- rep(0, length(noise))
curr_noise_x <- rnorm(nrow(data))
curr_noise_y <- rnorm(nrow(data))
for (i in 1 : length(noise)) {
    temp$x <- data$x * mult + curr_noise_x * noise[i]
    temp$y <- data$y * mult + curr_noise_y * noise[i]
    partial <- 0
    for (r in 1 : 12) {
        curr <- temp[month(temp$Date) %in% r, ]
        curr_r <- test(curr$x, curr$y, train_linear)[1]
        if (!is.na(curr_r) & curr_r == 1) {
            partial <- partial + 1
        }
    }

    acc[i] <- partial / 12.0
}

plot(noise, acc)
```

Notice that here the sd of the difference of the consecutive days is 4.326203. Notice that when noise is larger or equal to 8, approximately two sigma, the outputs are all NAs given zero accuracy.

# The Cause and Effect pairs

```{r allpairs, eval=FALSE, echo=TRUE}
pairs <- list.files("../pairs")
pairs <- grep("pair\\d{4}.txt", pairs, value = TRUE)
pair_coef <- data.frame(FileName = character(),
                        ForwardPValue = double(),
                        ForwardIntersection = double(),
                        ForwardBeta = double(),
                        BackwardPValue = double(),
                        BackwardIntersection = double(),
                        BackwardBeta = double(),
                        Result = factor())

for (pair in pairs) {
  xx <- read.delim(paste("../pairs/", pair, sep = ""), header = F, sep = "", stringsAsFactors = F)
  if (ncol(xx) != 2 || (length(unique(xx$V2)) == 2 && all(sort(unique(xx$V2)) == c(0, 1)))) {
    next
  }
  
  message("Pair Name: ", pair, " number of cols ", ncol(xx))
  
  yy <- test(as.numeric(xx$V1), as.numeric(xx$V2), reg_model = train_linear, detail = TRUE)
  
  curr_result <- factor("X->Y")
  if (is.na(yy[1])) {
      curr_result <- factor("Inconclusive")
  } else if (yy[1] == 0) {
      curr_result <- factor("Y->X")
  }
  
  curr <- list(FileName = pair,
               ForwardPValue = yy[2],
               ForwardIntersection = yy[3],
               ForwardBeta = yy[4],
               BackwardPValue = yy[5],
               BackwardIntersection = yy[6],
               BackwardBeta = yy[7],
               Result = curr_result)
  pair_coef <- rbind(pair_coef, curr, stringsAsFactors = FALSE)
}

write.csv(pair_coef, "../Allpair.csv", row.names = FALSE)
```


Let's get a closer look at pair43, we take samples of them and do tests to check if that helps.

```{r pair43}
pair <- "pair0043.txt"
pair_43 <- read.delim(paste("../pairs/", pair, sep = ""), header = F, sep = "", stringsAsFactors = F)

total <- 0

for (i in 1 : 10) {
  xx <- pair_43[sample(1 : (nrow(pair_43)), 1000), ]
  yy <- test(as.numeric(xx$V1), as.numeric(xx$V2), detail = F)
  curr_result <- factor("X->Y")
  if (is.na(yy[1])) {
      curr_result <- factor("Inconclusive")
  } else if (yy[1] == 0) {
      curr_result <- factor("Y->X")
  }
  
  message("The result is ", curr_result)
}

message("The whole accuracy is ", (total / 10.0))
```

Unfortunately we still got all wrong answers. Possible reasons could be this data set contains temperatue data from different places.

